---
title: "HW2 Prince George's County Crime"
---

In this assignment, we will be working with a dataset reporting the number of crimes by month in Prince George's County, Maryland, for every month in 2024. 

### Task 1: Load the Tidyverse

You will need to load the tidyverse library

```{r}
library(tidyverse)

```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/pg_crime_totals_2024.csv. You will create a dataframe called `pg_crime` that contains the data from this URL.

### Task 2: Load the Data and Describe It

```{r}
pg_crime <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/pg_crime_totals_2024.csv")
```
### Task 3: Describe the Data

You should examine that dataframe carefully by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the categories of incidents. From a reporting standpoint, are any of the categories more important than others? Why or why not? Do any of them make the job of summarizing the data more difficult? Why or why not?

Accidents are the most common incidents, with accidents involving impound close behind, which isn’t surprising given how frequent and normalized car accidents are. Theft from auto and auto theft also rank high in the data. From a reporting standpoint, accidents are especially important because they often affect more than just those directly involved, creating a broader public impact. Since they occur so often, journalists must decide which cases stand out as newsworthy. On the other hand, categories like residential robbery and homicide make summarizing the data more difficult because their relatively low numbers make them feel like outliers compared to the more frequent categories.

### Task 4: Calculating Averages

Calculate the total number of incidents per month for each category using `mutate()`, calling the new column `total_incidents`. Then calculate the average number of incidents per month for each category of crime, calling the new column `average_incidents`. Assign the result to a new dataframe called `pg_crime_averages`.

```{r}
pg_crime_averages <- pg_crime %>%
  group_by(category) %>%
  mutate(
    total_incidents = sum(Jan + Feb + Mar + Apr + May + Jun + Jul + Aug + Sep + Oct + Nov + Dec),
    average_incidents = mean(Jan + Feb + Mar + Apr + May + Jun + Jul + Aug + Sep + Oct + Nov + Dec))

  
```

### Task 5: Describe the Averages

Examine the new dataframe, `pg_crime_averages`, by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the average number of incidents per month for each category. Which categories have the highest average number of incidents per month? Which have the lowest? What does this tell you about the pattern of crime in Prince George's County? Are there categories that you have questions about? Why or why not?

Accidents have the highest average number of incidents per month, nearly average 100 incidents above any other category. Auto Stolen and Theft From Auto have some of the highest average number of incidents. Homicide and B&E Other have the lowest average number of incidents per month. Robbery Residential also has low numbers but does not yield an average.

This shows me that crime incidents involving automobiles tend to be the most frequent in Prince George’s County. Car accidents and theft from auto occur at a fairly consisten rate. 

I’m curious why Robbery Residential doesn’t yield an average. I assume it’s because some months have N/A rather than a number and therefore there isn’t enough information to calculate an average. I’m also interested in why Thefts are drastically higher in October. I wonder if it is related to Halloween. 


### Task 6: Testing a Theory

One theory about crime is that it is more likely to occur in the summer months. Test this theory by creating three new columns using `mutate()` that calculate the percentage change compared to the average for each category of crime for the months of June, July, and August. Call the new columns `june_change`, `july_change`, and `august_change`. Assign the result to a new dataframe called `summer_crime_averages`.

```{r}

summer_crime_averages <- pg_crime_averages %>%
mutate(
june_change = ((Jun - average_incidents) / average_incidents) * 100,
july_change = ((Jul - average_incidents) / average_incidents) * 100,
august_change = ((Aug - average_incidents) / average_incidents) * 100
)

```

### Task 7: Describe the Summer Crime Averages

Look at the `summer_crime_averages` dataframe by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the percentage change in the average number of incidents per month for each category of crime for the months of June, July, and August. Do you see any patterns? What does this tell you about the theory that crime is more likely to occur in the summer months?

Whether crime increased or decreased in the summer months seems to be mixed. Some categories, like homicide and auto stolen and recovered, experienced a positive percent change in at least two of the three summer months, meaning there were many more reported cases. However, other categories showed sharp decreases in summer months, like B&E other that had a decrease of 43% in June. I don’t think the data supports the story that crime is more likely to occur in the summer because it varies by category. 

### Task 8: What's the Story?

Write up a lede and nutgraf of a story focused on the crime data in Prince George's County. It could be from any of the dataframes you've created in this assignment. The lede should be a single sentence that summarizes the most important information in the data. The nutgraf should be a paragraph that provides more context and explains why the data is important.

It could be a spooky season for crime as theft cases reportedly heightened last October.

There were 363 thefts in October. The only month that got anywhere close to that number was January with 305 reported thefts. Halloween and shorter daylight hours could contribute to this increase. On Halloween, homeowners might leave their houses unlocked or pay less attention to their surroundings, making it easier for theft to occur. Shorter daylight hours would give more opportunities to thieves to act with less visibility outside. The heighten in the number of thefts for just one month would be an interesting topic to focus a story on.

### Task 9: Save Your Work and Put it on GitHub

Save this notebook, then commit it to your local repository. Push the changes to your GitHub repository and submit the URL for this file in ELMS.
