---
title: "HW3 WH Salaries"
---

In this assignment, we will be working with a dataset containing information about the salaries of White House staffers as of July 2024.

### Task 1: Load the Tidyverse

You will need to load the tidyverse library

```{r}
library(tidyverse)
install.packages("janitor")
library(janitor)
```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/biden_wh_salaries_2024.csv. You will create a dataframe called `wh_salaries` that contains the data from this URL.

### Task 2: Load the Data and Describe It

```{r}
wh_salaries <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/biden_wh_salaries_2024.csv")
```
### Task 3: Describe the Data

You should examine that dataframe carefully by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the categories of incidents. From a reporting standpoint, what stands out to you about this data? Are there any salary values that will cause problems for analysis?

ANSWER HERE
The salary values that are 0 could cause problems for my analysis because they are outliers.

### Task 4: Calculate the Mean and Median of the Salaries

Write code to calculate the mean and median of the salaries in the dataset using the `summarize()` function. You don't need to make a new dataframe, just display the results.

```{r}
    wh_salaries |> 
summarize(mean_wh_salaries = mean(salary),
            median_wh_salaries = median(salary),
)

```

### Task 5: Add in the minimum and maximum salaries

Take the code you just wrote and add in the minimum and maximum salaries (the functions for that are `min()` and `max()`). Your results should show four columns: mean_salary, median_salary, min_salary, max_salary.

```{r}

  wh_salaries |> 
summarize(mean_wh_salaries = mean(salary),
            median_wh_salaries = median(salary),
  max_wh_salaries = max(salary),
            min_wh_salaries = min(salary)
)
```

### Task 6: Describe the Results

Write a couple of sentences describing the results of your analysis. What do the mean and median salaries tell you about the data? What do the minimum and maximum salaries tell you? Are there any outliers in the data?

The mean and median salaries tell me that White House employees, on average, have high salaries. However, when we look at the min and max, we see how wide a variation there is between the highest and lowest salaries. It is a difference of over $250,000. Given this, outliers would be the min and max salaries. The maximum salary of $251,258 is significantly higher than the second highest of $191,900. Similarly, 11 people have salaries of $0, which could provide significant outliers in comparison to the rest of the data.

### Task 7: Story Idea

Based on the data, what is your best story idea? What is the next step you would take to develop that story idea?

I think a story could be written on the unpaid jobs. Why are there positions making no money when others are making upwards of $150,000? Are they getting money some other way? To develop this story idea, I would research the jobs that don't get paid and find out how much work the people in those positions are doing. I would also find out if historically those people or positions have been paid.
