---
title: "HW4 MD CityCrime"
---

In this assignment, we will be working with a dataset of total crimes reported by Maryland cities for two consecutive years. You'll load the data, calculate the violent and property crime rates for each city for both years, and then calculate the percentage change in the crime rates between the two years. You'll also write a short story based on your findings.

### Task 1: Load the Tidyverse and Janitor

You will need to load the tidyverse library and the janitor library

```{r}
library(tidyverse)
install.packages("janitor")
library(janitor)
```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_cities_crime_2019_2020.csv. You will create a dataframe called `md_cities` that contains the data from this URL.

### Task 2: Load the Data

```{r}
md_cities <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_cities_crime_2019_2020.csv")
```
### Task 3: Calculating Rates

Calculate violent and property crime rates for each city for both years using `mutate()`. You'll create four new columns: `violent_rate_2019`, `property_rate_2019`, `violent_rate_2020`, and `property_rate_2020`. You'll need to figure out the proper unit rate, given that there are varying populations. There's no perfect choice; choose the one that works best for you. Save the results to a new dataframe called `md_cities_rates`.

```{r}
 md_cities_rate <- md_cities |> mutate (
violent_rate_2019 = violent2019 / pop2019 * 1000,
property_rate_2019 = property2019 / pop2019 * 1000,
violent_rate_2020 = violent2020 / pop2020 * 1000,
property_rate_2020 = property2020/ pop2020 * 1000
)

```

### Task 4: Describe the Rates

Examine the new dataframe, `md_cities_rates`, by clicking on it in the Environment tab in RStudio and sorting by the rate columns. Write a couple of sentences about the results, comparing and contrasting the cities with the highest rates (both violent and property) in 2019 and 2020. Do you see any patterns? What does this tell you about crime in Maryland cities? What's the story here?

Crime rates are consistently high in Ocean City, according to the data. Ocean City had the highest or second-highest violent and property crime rate in 2019 and 2020. It was only beaten by Baltimore in violent crime rates in 2019.

I found this interesting because most people refer to Baltimore as the most dangerous place in Maryland, when this data shows that Ocean City is worse. This beach town attracts many tourists over the summer months, so I’m sure visitors would be interested in knowing how unsafe it can be. However, this influx of tourists each year might also be a contributing factor to the city’s higher crime rates.

There are some towns with crime rates of zero because they have had no recorded violent or property crime. This is partly due to their extremely small populations. For example, Barclay, a town in Queen Anne's County with no recorded crime, has a population of 167 in contrast to Ocean City’s population of 6926.



### Task 5: Calculate Percentage Change

Let's add two percentage change columns to your dataframe: one for violent crime rates and one for property crime rates. You'll calculate the percentage change between 2019 and 2020 for each city. Save the results to a new dataframe called `md_cities_change`.

```{r}

md_cities_change <- md_cities_rate |> mutate (
violent_change = (violent_rate_2020 - violent_rate_2019) / violent_rate_2019,
property_change = (property_rate_2020 - property_rate_2019) / property_rate_2019
)

```

### Task 7: Describe Percentage Changes

Look at the `md_cities_change` dataframe by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the percentage change results. What's more interesting here, the patterns or the outliers? What's the story in the data?

Most towns didn’t experience much of a change in violent crime rates from 2019 to 2020. However, Edmonston and Poolesville both had a change of at least a 6% increase. Similarly, Crisfield has a 3% increase in property crime, and most places stayed fairly consistent. I think the pattern is more interesting since the crime rate in most towns differed only slightly from 2019 to 2020. I think the story could be about the consistency in crime and how it’s not decreasing.

### Task 8: Tell the Best Story

Write up a lede and nutgraf of a story with the _best_ story idea you've found in this assignment. The lede should be a single sentence that summarizes the most important information in the data. The nutgraf should be a paragraph that provides more context and explains why the data is important.

Is Ocean City the most dangerous city? High violent and property crime rates might cause tourists to reconsider their vacation spot.

In 2019, Ocean City had the highest property crime rate, and it was only barely behind Baltimore with the highest violent crime rate that year. In 2020, Ocean City led both categories with the highest crime rates. This might cause problems for tourists who travel to the town during breaks and summers to visit the beach.


### Task 9: Save Your Work and Put it on GitHub

Save this notebook, then commit it to your local repository. Push the changes to your GitHub repository and submit the URL for this file in ELMS.
